Synthesizing Quotient Lenses

### 2018-03-13 papers/QuotientLenses/main.tex
pass through section 4

% FOR Regular Expression names
\newcommand{\re}[1]{\ensuremath{\mathtt{#1}}}
\newcommand{\kw}[1]{\ensuremath{\mathit{#1}}}
\newcommand{\codefont}[1]{\ensuremath{\mathsf{#1}}}
\newcommand{\project}[2]{\ensuremath{\kw{project} \; #1 \mapsto #2}}
\newcommand{\squash}[3]{\ensuremath{\kw{squash} \; #1 \rightarrow #2\; \kw{using} \; #3}}
\newcommand{\perm}[2]{\ensuremath{\kw{perm}\; (#1)\; \kw{with}\; #2}}
\newcommand{\normalize}[3]{\ensuremath{\kw{normalize} \; (#1, #2, #3)}}
\newcommand{\eqrel}[1]{\ensuremath{\equiv_{#1}}}
\newcommand{\sep}{\ensuremath{\ | \ }}
\newcommand{\canonize}{\ensuremath{\kw{canonize}}}
\newcommand{\bibtex}{\textsc{Bib}\TeX{}}
\newcommand{\get}{\ensuremath{\kw{get}}}
\newcommand{\semicolon}{\ensuremath{\; ; \;}}
\newcommand{\lput}{\ensuremath{\kw{put}}}
\newcommand{\create}{\ensuremath{\kw{create}}}
\newcommand{\const}{\ensuremath{\kw{const}}}
\newcommand{\swap}{\ensuremath{\kw{swap}}}
\newcommand{\id}{\ensuremath{\kw{id}}}
\newcommand{\lquot}{\ensuremath{\kw{lquot}}}
\newcommand{\rquot}{\ensuremath{\kw{rquot}}}
\newcommand{\lift}{\ensuremath{\kw{lift}}}

@@

\begin{document}
\title{Synthesizing Quotient Lenses}
\begin{abstract}
{\em Quotient lenses} are bidirectional transformations whose correctness
laws are ``loosened'' by specified equivalence relations, allowing
inessential details in concrete data formats to be suppressed. 
For example, a programmer could use a quotient lens to define 
a transformation that ignores the order of fields in XML data, so
that two XML files with the same fields but in different orders would be
considered the same, allowing a single, simple lens to handle them both.

This paper simplifies the task of programming quotient lenses with
three connected innovations.
First, we introduce {\em quotient regular expressions} (QREs), to
easily mark aspects of data formats that are
inessential.  A QRE is an augmented regular expression that
simultaneously defines a regular language and an equivalence
relation on this language.  Variations within the same
equivalence class are deemed inessential. 
Next, we introduce {\em QRE lenses}---lenses mapping
between QREs.
Our key technical result is a proof that every QRE lens can be
transformed into a QRE lens that has the form $c_1 \cdot \ell \cdot
c_2$, where $c_1$ and $c_2$ are {\em canonizers} 
and $\ell$ is a bijective transformation.  Intuitively,
the canonizers normalize data by picking a
canonical representative of the relevant equivalence relations. 
This theorem means we can push all normalization to the outer edges of a
bidirectional transformation, leaving in the middle a pure bijective
transformation.  
Finally,
we leverage this normalization theorem to {\em synthesize} QRE lenses from a pair
of QREs and example input-output pairs, reusing earlier work on
synthesizing plain bijective lenses.
We have implemented QREs and QRE lens synthesis 
in the bidirectional programming language Boomerang.
We evaluate the effectiveness of our approach by using it to
synthesize QRE lenses between various real-world data formats in
the {\tt data.gov} database.\bcp{The very last bit is inaccurate now.}
\end{abstract}

@@

\section{Quotient Regular Expressions}
\label{QRE}
A Quotient Regular Expression (or QRE) is a regular expression $R$ augmented
with syntax that expresses an equivalence relation on the language of $R$. This
section describes in further detail the set of QRE combinators that we described
in \cref{sec:example}, but first we discuss the \kw{normalize} combinator which
enjoys some special properties.

\subsection{The \kw{normalize} Combinator}
The \kw{normalize} combinator is special in two main ways, the first of
which has to do with the premises in its inference rule:
\begin{prooftree}
\AxiomC{$\mathcal{L}(R') \subseteq \mathcal{L}(R)$}
\AxiomC{$f : \mathcal{L}(R) \longrightarrow \mathcal{L}(R')$}
\AxiomC{$f$ is surjective}
\AxiomC{$f = f^2$}
\RightLabel{(Normalize)}
\QuaternaryInfC{$\wf{\normalize{R}{R'}{f}}$}
\end{prooftree}
In order for the {\em normalize} QRE to be well-formed, then canonizing function
$f$ used by this combinator must be surjective and idempotent. Unfortunately,
verifying that a function satisfies these properties is in general undecidable.
Nonetheless, we allow a programmer to use normalize at their own risk without
checking these side conditions.

We do this firstly because the {\em normalize} combinator allows one to express
equivalence relations that are either difficult to define, or that cannot be
defined using the other combinators. Secondly, the {\em normalize} canonizer
enjoys the property that each of the other combinators $Q$ is semantically
equivalent to $\normalize{W(Q)}{K(Q)}{f}$ for some surjective, idempotent
function $f : \mathcal{L}(W(Q)) \longrightarrow \mathcal{L}(K(Q))$; this is the
second property that distinguishes the {\em normalize} combinator from the other
QREs. The {\em normalize} combinator hence gives a sufficient condition for
potential QREs to be ``well-behaved'', particularly with respect to the
strategy for synthesizing quotient lenses which we shall introduce in
Section~\ref{synth}.

Having discussed the special properties of the \kw{normalize} combinator, we now
give the syntax and semantics of the other QRE combinators.
\subsection{Syntax and Semantics of QREs}
Formally, the language of Quotient Regular Expressions (QREs) is given by the
following grammar:
\begin{align*}
Q := \; &\id(R) \sep \project{R}{s} \sep \squash{Q_1}{Q_2}{f} \sep
\perm{Q_1, \ldots, Q_n}{Q} \\
& | \; \normalize{R}{R'}{f} \sep Q_1 \semicolon Q_2 \sep Q_1 \cdot Q_2 \sep (Q_1
\sep Q_2) \sep Q^*,
\end{align*}
where $Q$ ranges over QREs, $R$ ranges over regular expressions, $f$ ranges over
functions between regular languages, and $s$ ranges over character strings.

Using the conventional notation that $\mathcal{L}(R)$ is the language accepted
by the regular expression $R$, then each QRE $Q$ enables us to express
\begin{enumerate}
  \item a regular expression $W(Q)$ (the ``whole'' of $Q$),
  \item an equivalence relation $\eqrel{Q}$ on $\mathcal{L}(W(Q))$,
  \item a regular expression $K(Q)$ (the ``kernel'' of $Q$)
  such that $\mathcal{L}(K(Q))$ forms a complete set of representatives for
  $\eqrel{Q}$, and
  \item a ``canonizing'' function $\canonize(Q):\mathcal{L}(W(Q))
  \longrightarrow \mathcal{L}(K(Q))$ which given any $w \in \mathcal{L}(W(Q))$,
  computes $\canonize(Q)(w)$ as the unique $k$ in $\mathcal{L}(K(Q))$ such
  that $k$ is equivalent to $w$ mod $\eqrel{Q}$
\end{enumerate}
Figures~\ref{fig:wk} gives the inductive definitions for the whole and
kernel languages $W(Q)$ and $K(Q)$ of a QRE.
\begin{figure}[t]
\centering
\[
\begin{array}{l@{\quad}l@{\quad}l}

Q & W(Q) & K(Q) \\ \hline
R & R & R \\
\project{R}{s} & R & s \\
\squash{Q}{Q_1}{f} & W(Q) \sep W(Q_1) & K(Q_1) \\
\normalize{R}{R'}{f} & R & R' \\
Q_1 \; ; \; Q_2 & W(Q_1) & K(Q_2) \\
Q_1 \cdot Q_2 & W(Q_1) \cdot W(Q_2) & K(Q_1) \cdot K(Q_2) \\
Q_1 \sep Q_2 & W(Q_1) \sep W(Q_2) & K(Q_1) \sep K(Q_2) \\
Q^* & W(Q)^* & K(Q)^* \\
\end{array}
\]
\[
\begin{array}{r@{\quad}l}
W( \perm{Q_1, \ldots, Q_n}{Q} ) = &
\bigcup \limits_{\sigma \in S_n} W(c_{\sigma(1)}) \cdot W(Q) \cdot \ldots \cdot
W(Q) \cdot W(c_{\sigma(n)})\\
K( \perm{Q_1, \ldots, Q_n}{Q} ) = & K(Q_1) \cdot K(Q) \cdot \ldots \cdot K(Q)
\cdot K(Q_n)
\end{array}
\]
\caption{Whole and Kernel Regular Expressions}
\label{fig:wk}
\end{figure}
The \textit{squash} and \textit{permutation} combinators give the two
interesting definitions. If $Q = \squash{Q_1}{Q_2}{f}$, then the whole language
of $Q$ is $W(Q_1) \sep W(Q_2)$. This is because the \textit{squash} combinator
takes the whole language $W(Q_1)$ of $Q_1$, the whole language $W(Q_2)$ of $Q_2$
and a function $f : \mathcal{L}(W(Q_1)) \longrightarrow \mathcal{L}(W(Q_2))$,
maps $\mathcal{L}(W(Q_1))$ into $\mathcal{L}(W(Q_2))$ using $f$ and then
canonizes $W(Q_2)$ into $K(Q_2)$ using $\canonize(Q_2)$.

For the $\perm{Q_1, \ldots, Q_n}{Q}$ combinator, the whole language is the
union of languages of the form $W(c_{\sigma(1)}) \cdot W(Q) \cdot \ldots \cdot
W(Q) \cdot W(c_{\sigma(n)})$ for any permutation $\sigma$ in $S_n$, where
$S_n$ is the set of all permutations of the numbers $1$ to $n$. This is because
the \textit{permutation} combinator allows for the string to match any
permutation the $Q_i$'s while also taking into account the separator $Q$ in
between each of the $Q_i$'s. The kernel of the \textit{permutation} combinator
is the language $K(Q_1) \cdot K(Q) \cdot \ldots \cdot K(Q) \cdot K(Q_n)$,
because the canonical permutation is the identity permuation, with each of the
parts of the string that match $Q_i$ and $Q$ canonized into $K(Q_i)$ and $K(Q)$
respectively.

Figure~\ref{fig:canonizers} gives the inductive definitions of the
$\canonize{}$ function for each QRE.
\begin{figure}[t]
\begin{center}
\[
\begin{array}{l@\quad l @\quad l}
\canonize(\id(R)) &=& id_{\mathcal{L}(R)} \\
\canonize(\project{R}{s})(w) &=& s\\
\canonize(\squash{Q_1}{Q_2}{f})(w) &=&
\begin{cases}
\canonize(Q_1)(f(w)) & \text{if } w \in \mathcal{L}(W(Q_1))\\
\canonize(Q_2)(w) & \text{otherwise}
\end{cases}\\
\canonize(\normalize{R}{R'}{f}) &=& f\\
\canonize(Q_1 \; ; \; Q_2) &=& \canonize(Q_2) \circ \canonize(Q_1)\\
\canonize(Q_1 \cdot Q_2) &=& \canonize(Q_1) \cdot \canonize(Q_2)\\
\canonize(Q_1 \sep Q_2)(w) &=&
\begin{cases}
\canonize(Q_1)(w) & \text{if } w \in \mathcal{L}(W(Q_1))\\
\canonize(Q_2)(w) & \text{if } w \in \mathcal{L}(W(Q_2))\\
\end{cases}\\
\canonize(Q^*) &=& \canonize(Q)^* \\
\end{array}
\]
\end{center}
\begin{align*}
&\canonize(\perm{Q_1, \ldots, Q_n}{Q})(r_{\sigma(1)}
\cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)})\\
&= (\canonize(Q_1)(r_1)) \cdot (\canonize(Q)(s_1)) \cdot \ldots \cdot
(\canonize(Q_n)(r_n))
\end{align*}
\caption{QRE canonizers}
\label{fig:canonizers}
\end{figure}
Again, the \textit{permutation} combinator gives the most interesting
definition,
\begin{align*}\canonize(\perm{Q_1, \ldots, Q_n}{Q})(r_{\sigma(1)}
\cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)}) \\
= (\canonize(Q_1)(r_1)) \cdot (\canonize(Q)(s_1)) \cdot \ldots \cdot
(\canonize(Q_n)(r_n))
\end{align*}
\noindent which, as we have explained, is defined so that the strings that match
$Q_i, Q$ are placed in the canonical permutation which is the identity
permutation before $\canonize(Q_i)$ and $\canonize(Q)$ are applied
respectively.

@@

The most complicated inference rule is the (Perm) rule for the
\textit{permutation} combinator. The second hypothesis for the
\textit{permutation} rule says that for any two different permutations $\sigma$
and $\theta$, the languages $W(c_{\sigma(1)}) \cdot W(Q) \cdot \ldots \cdot
W(c_{\sigma(n)})$ and $W(c_{\theta(1)}) \cdot W(Q) \cdot \ldots \cdot
W(c_{\theta(n)})$ must be disjoint. This is because any string that matches the
\textit{permutation} combinator matches the language $W(c_{\theta(1)}) \cdot
W(Q) \cdot \ldots \cdot W(c_{\theta(n)})$ for some permutation $\theta$, so we
require that these regular expressions be disjoint.
%
The third hypothesis says that the regular expression $K(Q_1) \cdot K(Q)
\cdot \ldots \cdot K(Q) \cdot K(Q_n)$ must be unambiguous. This is because the
underlying lens that maps to or from a \textit{permutation} QRE operates on
the language $K(Q_1) \cdot K(Q) \cdot \ldots \cdot K(Q) \cdot K(Q_n)$ which is
the kernel of the \textit{permutation} QRE. As we mentioned, the underlying
lens will require that the source and target regular expressions of the lens be
unambiguous so that the lens can match strings uniquely.

>>>>

% FOR Regular Expression names
\newcommand{\re}[1]{\ensuremath{\mathtt{#1}}}
\newcommand{\codefont}[1]{\ensuremath{\mathsf{#1}}}
\newcommand{\kw}[1]{\codefont{#1}}
\newcommand{\project}[2]{\ensuremath{\kw{project} \; #1 \mapsto #2}}
\newcommand{\squash}[3]{\ensuremath{\kw{squash} \; #1 \rightarrow #2\; \kw{using} \; #3}}
\newcommand{\perm}[2]{\ensuremath{\kw{perm}\; (#1)\; \kw{with}\; #2}}
\newcommand{\normalize}[3]{\ensuremath{\kw{normalize}(#1, #2, #3)}}
\newcommand{\eqrel}[1]{\ensuremath{\equiv_{#1}}}
\newcommand{\sep}{\ensuremath{\ | \ }}
\newcommand{\canonize}{\ensuremath{\kw{canonize}}}
\newcommand{\bibtex}{\textsc{Bib}\TeX{}}
\newcommand{\get}{\ensuremath{\kw{get}}}
\newcommand{\semicolon}{\ensuremath{\; ; \;}}
\newcommand{\lput}{\ensuremath{\kw{put}}}
\newcommand{\create}{\ensuremath{\kw{create}}}
\newcommand{\const}{\ensuremath{\kw{const}}}
\newcommand{\swap}{\ensuremath{\kw{swap}}}
\newcommand{\id}{\ensuremath{\kw{id}}}
\newcommand{\lquot}{\ensuremath{\kw{lquot}}}
\newcommand{\rquot}{\ensuremath{\kw{rquot}}}
\newcommand{\lift}{\ensuremath{\kw{lift}}}

@@

\begin{document}
\title{Synthesizing Quotient Lenses}
\begin{abstract}
{\em Quotient lenses} are bidirectional transformations that respect
specified equivalence relations with the goal of 
appropriately handling inessential details in data formats. 
For example, a programmer could use a quotient lens to define 
a transformation that ignores the order of fields in XML data, so
that two XML files with the same fields but in different orders would be
considered the same. 
This paper simplifies the task of programming quotient lenses in three
steps.
First, we introduce {\em quotient regular expressions} (QREs) to
make it easy to express which portions of a data format are
inessential.  A QRE is an augmented regular expression that
simultaneously defines a regular expression and an equivalence
relation over matching strings.  Variations within the same
equivalence class are deemed inessential. 
Next, we introduce {\em QRE lenses}, which are lenses that map
between QREs.
Our key technical result is a proof that every QRE lens can be
transformed into a QRE lens that has the form $c_1 \cdot \ell \cdot
c_2$, where $c_1$ and $c_2$ are {\em canonizers} 
and $\ell$ is a bijective transformation.  Intuitively,
the canonizers normalize the data  by picking a
canonical representative of the relevant equivalence relations. 
This theorem means we can push all the normalization to the edges of a
bidirectional transformation, leaving in the middle a pure bijective
transformation.  
Finally,
we levarage this insight to {\em synthesize} QRE lenses from a pair
of QREs and example input-output pairs by reusing earlier work on
synthesizing bijective lenses.
We have implemented QREs and QRE lens synthesis 
in the bidirectional programming language Boomerang.
We evaluate the effectiveness of our approach by using it to
synthesize QRE lenses between various real-world data formats in
the {\tt data.gov} database.

\end{abstract}

@@

\section{Quotient Regular Expressions}
\label{QRE}
A Quotient Regular Expression (or QRE) is a regular expression $R$ augmented
with syntax that expresses an equivalence relation on the language of $R$. This
section describes in further detail the set of QRE combinators that we described
in \cref{sec:example}.\saz{Why does the ``cref'' macro not capitalize Section?}

\subsection{Syntax of Semantics of QREs}
Formally, the language of Quotient Regular Expressions (QREs) is given by the
following grammar:
\newcommand{\bsep}{\ \ \sep{} \ \ }
\begin{align*}
Q := \; & \normalize{R}{R'}{f} \bsep{} \id(R) \bsep \project{R}{s} \bsep \squash{Q_1}{Q_2}{f} \\
 | \; & \perm{Q_1, \ldots, Q_n}{Q} \bsep Q_1 \semicolon Q_2 \bsep Q_1 \cdot Q_2 \bsep (Q_1
\sep Q_2) \bsep Q^*,
\end{align*}
where $Q$ ranges over QREs, $R$ ranges over regular expressions, $f$ ranges over
functions between regular languages, and $s$ ranges over character strings.

Using the conventional notation that $\mathcal{L}(R)$ is the language accepted
by the regular expression $R$, each QRE $Q$ enables us to express four semantic objects

\begin{tabular}{rl}
  $W(Q)$  & a regular expression (the ``whole'' of $Q$),\\
  $\eqrel{Q}$  & an equivalence relation on $\mathcal{L}(W(Q))$,\\
  $K(Q)$ & a regular expression (the ``kernel'' of $Q$) such that
           $\mathcal{L}(K(Q))$ forms a complete \\
             & set of representatives for $\eqrel{Q}$, and \\
  $\canonize(Q)$ & 
  a ``canonizing'' function that, given any $w \in \mathcal{L}(W(Q))$,
  yields $\canonize(Q)(w)$ \\ & as the unique $k$ in $\mathcal{L}(K(Q))$ such
  that $k$ is equivalent to $w$ mod $\eqrel{Q}$
\end{tabular}
  % of type $\mathcal{L}(W(Q)) \longrightarrow \mathcal{L}(K(Q))$  which

The well-formedness constraints for QREs ensure that these four semantic
components fit together to form a coherent quotient $W(Q)/\eqrel{Q}$ whose
equivalence classes are determined by \canonize(Q).

\subsection{The \kw{normalize} Combinator}

The relationship among the semantic objects of a QRE can be understood in terms
of the \normalize{R}{S}{f} combinator, which exposes each of these pieces
explicitly.  Its whole language is just $R$, its kernel language is just
$S$, and $\canonize(\normalize{R}{S}{f}) = f$.  Its equivalence relation is
determined by the fibres of $f$, so we have
$s_1 \equiv s_2 \Leftrightarrow f(s_1) = f(s_2)$.

These components form a coherent quotient language when the canonization
function $f$ is surjective and idempotent (intuitively, $f$ picks out a unique
representative for each equivalence class).  We also require that the kernel
language be a subset of the whole language, which enables QRE composition.
These considerations lead to the following well-formedness rule.

\begin{prooftree}
\AxiomC{$\mathcal{L}(S) \subseteq \mathcal{L}(R)$}
\AxiomC{$f : \mathcal{L}(R) \longrightarrow \mathcal{L}(S)$}
\AxiomC{$f$ is surjective}
\AxiomC{$f = f^2$}
\RightLabel{(Normalize)}
\QuaternaryInfC{$\wf{\normalize{R}{R'}{f}}$}
\end{prooftree}

Semantically, the \kw{normalize} QRE is universal---each of the other
combinators $Q$ is equivalent to $\normalize{W(Q)}{K(Q)}{f}$ for some
surjective, idempotent function
$f : \mathcal{L}(W(Q)) \longrightarrow \mathcal{L}(K(Q))$.   However,
although \kw{normalize} is powerful, verifying that a canonization function $f$
is surjective and idempotent is in general undecidable.  This means that a
programmer wishing to use \kw{normalize} must discharge strong proof
obligations, which is cumbersome in practice.\footnote{In our implementation we allow a programmer to use \kw{normalize} at their
own risk without checking these side conditions as an ``escape hatch'' for the
case when other QRE combinators are insufficient.}

The remaining QRE combinators provide simpler, more compositional ways of
building canonizers.  Nevertheless, the \kw{normalize} combinator gives a
sufficient condition for potential QREs to be well-behaved, particularly with
respect to the strategy for synthesizing quotient lenses which we shall
introduce in Section~\ref{synth}, so we use it to guide the design of our QRE
combinators, as we see next.

% ; this is the second
% property that distinguishes the \kw{normalize} combinator from the other QREs.
% The \kw{normalize} combinator is special in two main ways, the first of 
% which has to do with the premises in its inference rule:

\subsection{QRE Combinator Semantics}

Figure~\ref{fig:wk} gives the inductive definitions of the whole and
kernel languages $W(Q)$ and $K(Q)$ for all of the QRE forms.
\begin{figure}[t]
\centering
\[
\begin{array}{l@{\quad}l@{\quad}l}

Q & W(Q) & K(Q) \\ \hline
R & R & R \\
\project{R}{s} & R & s \\
\squash{Q}{Q_1}{f} & W(Q) \sep W(Q_1) & K(Q_1) \\
\normalize{R}{R'}{f} & R & R' \\
Q_1 \; ; \; Q_2 & W(Q_1) & K(Q_2) \\
Q_1 \cdot Q_2 & W(Q_1) \cdot W(Q_2) & K(Q_1) \cdot K(Q_2) \\
Q_1 \sep Q_2 & W(Q_1) \sep W(Q_2) & K(Q_1) \sep K(Q_2) \\
Q^* & W(Q)^* & K(Q)^* \\
\end{array}
\]
\[
\begin{array}{r@{\quad}l}
W( \perm{Q_1, \ldots, Q_n}{Q} ) = &
\bigcup \limits_{\sigma \in S_n} W(c_{\sigma(1)}) \cdot W(Q) \cdot \ldots \cdot
W(Q) \cdot W(c_{\sigma(n)})\\
K( \perm{Q_1, \ldots, Q_n}{Q} ) = & K(Q_1) \cdot K(Q) \cdot \ldots \cdot K(Q)
\cdot K(Q_n)
\end{array}
\]
\caption{Whole and Kernel Regular Expressions}
\label{fig:wk}
\end{figure}
The \kw{squash} and \kw{permutation} combinators have the two most
interesting definitions. If $Q = \squash{Q_1}{Q_2}{f}$, then the whole language
of $Q$ is $W(Q_1) \sep W(Q_2)$. This is because the \kw{squash} combinator
takes the whole language $W(Q_1)$ of $Q_1$, the whole language $W(Q_2)$ of $Q_2$
and a function $f : \mathcal{L}(W(Q_1)) \longrightarrow \mathcal{L}(W(Q_2))$,
maps $\mathcal{L}(W(Q_1))$ into $\mathcal{L}(W(Q_2))$ using $f$ and then
canonizes $W(Q_2)$ into $K(Q_2)$ using $\canonize(Q_2)$.

For the $\perm{Q_1, \ldots, Q_n}{Q}$ combinator, the whole language is the
union of languages of the form $W(c_{\sigma(1)}) \cdot W(Q) \cdot \ldots \cdot
W(Q) \cdot W(c_{\sigma(n)})$ for any permutation $\sigma$ in $S_n$, where
$S_n$ is the set of all permutations of the numbers $1$ to $n$. This is because
the \kw{permutation} combinator allows for the string to match any
permutation the $Q_i$'s while also taking into account the separator $Q$ in
between each of the $Q_i$'s. The kernel of the \kw{permutation} combinator
is the language $K(Q_1) \cdot K(Q) \cdot \ldots \cdot K(Q) \cdot K(Q_n)$,
because the canonical permutation is the identity permuation, with each of the
parts of the string that match $Q_i$ and $Q$ canonized into $K(Q_i)$ and $K(Q)$
respectively.

Figure~\ref{fig:canonizers} gives the inductive definitions of the
$\canonize{}$ function for each QRE.
\begin{figure}[t]
\begin{center}
\[
\begin{array}{l@\quad l @\quad l}
\canonize(\id(R)) &=& id_{\mathcal{L}(R)} \\
\canonize(\project{R}{s})(w) &=& s\\
\canonize(\squash{Q_1}{Q_2}{f})(w) &=&
\begin{cases}
\canonize(Q_1)(f(w)) & \text{if } w \in \mathcal{L}(W(Q_1))\\
\canonize(Q_2)(w) & \text{otherwise}
\end{cases}\\
\canonize(\normalize{R}{R'}{f}) &=& f\\
\canonize(Q_1 \; ; \; Q_2) &=& \canonize(Q_2) \circ \canonize(Q_1)\\
\canonize(Q_1 \cdot Q_2) &=& \canonize(Q_1) \cdot \canonize(Q_2)\\
\canonize(Q_1 \sep Q_2)(w) &=&
\begin{cases}
\canonize(Q_1)(w) & \text{if } w \in \mathcal{L}(W(Q_1))\\
\canonize(Q_2)(w) & \text{if } w \in \mathcal{L}(W(Q_2))\\
\end{cases}\\
\canonize(Q^*) &=& \canonize(Q)^* \\
\end{array}
\]
\end{center}
\begin{align*}
&\canonize(\perm{Q_1, \ldots, Q_n}{Q})(r_{\sigma(1)}
\cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)})\\
&= (\canonize(Q_1)(r_1)) \cdot (\canonize(Q)(s_1)) \cdot \ldots \cdot
(\canonize(Q_n)(r_n))
\end{align*}
\caption{QRE canonizers}
\label{fig:canonizers}
\end{figure}
Again, the \kw{permutation} combinator gives the most interesting
definition,
\begin{align*}\canonize(\perm{Q_1, \ldots, Q_n}{Q})(r_{\sigma(1)}
\cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)}) \\
= (\canonize(Q_1)(r_1)) \cdot (\canonize(Q)(s_1)) \cdot \ldots \cdot
(\canonize(Q_n)(r_n))
\end{align*}
\noindent which, as we have explained, is defined so that the strings that match
$Q_i, Q$ are placed in the canonical permutation which is the identity
permutation before $\canonize(Q_i)$ and $\canonize(Q)$ are applied
respectively.

@@

The most complicated inference rule is the (Perm) rule for the
\kw{permutation} combinator. The second hypothesis for the
Perm rule says that for any two different permutations $\sigma$
and $\theta$, the languages $W(c_{\sigma(1)}) \cdot W(Q) \cdot \ldots \cdot
W(c_{\sigma(n)})$ and $W(c_{\theta(1)}) \cdot W(Q) \cdot \ldots \cdot
W(c_{\theta(n)})$ must be disjoint. This is because any string that matches the
\kw{permutation} combinator matches the language $W(c_{\theta(1)}) \cdot
W(Q) \cdot \ldots \cdot W(c_{\theta(n)})$ for some permutation $\theta$, so we
require that these regular expressions be disjoint.
%
The third hypothesis says that the regular expression $K(Q_1) \cdot K(Q)
\cdot \ldots \cdot K(Q) \cdot K(Q_n)$ must be unambiguous. This is because the
underlying lens that maps to or from a \kw{permutation} QRE operates on
the language $K(Q_1) \cdot K(Q) \cdot \ldots \cdot K(Q) \cdot K(Q_n)$ which is
the kernel of the \kw{permutation} QRE. As we mentioned, the underlying
lens will require that the source and target regular expressions of the lens be
unambiguous so that the lens can match strings uniquely.
